### 2021-12-19 newvariant.tex
some rewrites

Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
Here, we challenge a commonly-made assumption in estimating the transmission advantage of SARS-CoV-2 variants and argue that holding relative strength (rather than relative speed) constant is generally more appropriate.
In doing so, we use the generation-interval-based framework to compare how relative strength and speed change over time under two idealized interventions:
a constant-strength intervention like idealized vaccination and social distancing, which reduces transmission rates by a constant amount, and a constant-speed intervention like idealized contact tracing, which isolates infected individuals at a constant rate.
Using the generation-interval-based framework further allows us to account for the possibility that new SARS-CoV-2 variants, such as the Alpha and Delta variants, may have different generation intervals from prior dominant lineages. 
Differences in the generation-interval distributions can affect the effectiveness of these interventions as well as the realized transmission advantages of new variants.
Neglecting these differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.

@@

Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
The generation interval is also distinctly different from other ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.

@@

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict relative speed \citep{davies2021estimated,di2021impact}.

While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the pathogen and its control \citep{doi:10.1098/rspb.2020.1556}.
To illustrate the differences in implications for holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed in this paper.
Given a pre-intervention kernel $K(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ regardless of age of infection ($K(\tau)/\theta$)---in this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
On the other hand, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$: $K(\tau) \exp(-\phi \tau)$---in this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to prevent onward transmission \citep{doi:10.1098/rspb.2020.1556}.

@@

Most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but Kissler \textit{et al.} \citep{kissler2021densely} suggested that the Alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type.
Longer duration of infection suggests that the mean generation interval of the Alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
In contrast, some studies have suggested that the more recent Delta variant may have a shorter generation interval \citep{li2021viral,zhang2021transmission}.
Modeling studies have also considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
However, linking strength and speed is complicated given that the generation-interval distribution depends on many factors including behavior:
for example, self-isolation after symptom onset will lead to shorter generation intervals.

@@

Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not change through time \citep{fraser2007estimating}.
In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events \citep{champredon2015intrinsic}.
Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of constant-strength changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}---even though the instantaneous generation-interval distribution remains unchanged in this scenario.
While the instantaneous generation-interval distribution can change under speed-like changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that control strategies against its spread have been primarily strength-like, including social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
Indeed, many dynamical models of SARS-CoV-2 infections have solely relied on constant-strength changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution (e.g., \citep{flaxman2020Rt,gostic2020practical,brauner2021inferring}).
Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of constant-strength changes on the inference of dynamics of new SARS-CoV-2 variants.
We revisit these ideas in Section 6 and compare the effects of constant-speed interventions with those of constant-strength interventions.

@@

While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
Focusing on proportions is advantageous, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
\begin{align}
p(t) &= \frac{\iv(t_0) \exp(\rv t)}{\iw(t_0) \exp(\rw t) + \iv(t_0) \exp(\rv t)},
\\ &= \frac{1}{1 + \left(\iw(t_0)/\iv(t_0)\right) \exp(-\delta t)},
\end{align}
where the relative speed of the variant $\delta$ can be estimated as the slope of the log odds of $p$  vs.~time.
When more than two strains are co-circulating, the picture is more complicated \citep{campbell2021increased}; we focus here on comparing two strains at a time.

@@

% We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
Inference of the relative strength $\rho = \Rv/\Rw$ from the observed relative speed $\delta$ depends on assumptions about the generation-interval distributions of both strains.
Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
\begin{equation}
\rho = \left(\frac{1 + \kappa (\rw + \delta) \Gv}{1 + \kappa \rw \Gw}\right)^{1/\kappa}.
\end{equation}
Therefore, the relative strength $\rho$ depends not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population (\rw)---
some analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case, $\rho = \exp(\delta \Gw)$ when $\Gv=\Gw$).

@@

We do not generally expect the relative speed $\delta$ to remain constant if other factors governing epidemic spread are changing.
Instead, many biological mechanisms appear compatible with assuming a constant value of relative strength $\rho$ over changing conditions.
For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}; 
as we show in this section, these changes will imply a change in relative speed $\delta$.

We thus investigate how $\delta$ is expected to change with \Rw when $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
\begin{equation}
\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \Gv} - \frac{\Rw^{\kappa} - 1}{\kappa \Gw}.
\end{equation}
As our baseline scenario we assume $\rho = 1.61$ (and therefore, we always have $\Rw < \Rv$), which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.

@@

\fref{relspeed} also shows that when $\rho$ is fixed, relative speed depends on underlying epidemiological conditions---specifically, the absolute strength of the two strains. 
For example, even when the generation-interval distributions are identical ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields a relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
Therefore, characterizing the spread of variants by assuming constant relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant} and using the resulting value of $\delta$) should be done with care.

@@

Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the Alpha variant by \cite{volz2021transmission}.
Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}; 
therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission \citep{gostic2020practical}.
The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.

@@

We consider two types of intervention:
an intervention of constant strength (\fref{strengthspeed}A,C,E), which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed (\fref{strengthspeed}B,D,F), which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively \citep{doi:10.1098/rspb.2020.1556}.
We consider constant-strength and -speed interventions calibrated to reduce $\Rv$ to 0.9 under the assumption that the variant generation intervals match those of the wild type (\fref{strengthspeed}C--D).
While both interventions are equally effective on the strength scale (that is, $\Ry{post}=\int  K_{\mathrm{post}}(\tau) \dtau = 0.9$), they have different dynamical implications.
The constant-strength intervention affects transmission equally throughout the course of infection, whereas the constant-speed intervention has greater impact on transmission that occurs later in infection;
as a result, the constant-speed intervention reduces the post-intervention mean generation interval (\fref{strengthspeed}B) and leads to (slightly) faster exponential decay (therefore, lower $\ry{post}$).

@@

We explored how the generation-interval distribution shapes the link between relative strength and speed.
While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, using the generation-interval-based framework provides clear explanations for an important, but under-appreciated concept: the relative strength and speed of new variants also depend on epidemic dynamics of the prior dominant lineages as well as the type of control measures (i.e., strength- or speed-like) that limit their spread.
For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain---this challenges the commonly-made assumption that relative speed of new variants remains constant over time.
Differences in the generation-interval distributions further exaggerate this effect.

Therefore, neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data.
These biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant.
Differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
If the variant has longer generation intervals than the wild type, speed-like interventions will be relatively more effective than naive estimates would suggest. 
Conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals.

As new variants of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants (and the corresponding uncertainty).
If new variants have different infectiousness profiles \citep{kissler2021densely,li2021viral,zhang2021transmission}, and therefore generation-interval distributions, current estimates that do not account for different generation-interval distributions may be systematically biased.
% If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.

There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for the Alpha variant.
Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the Alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences:
For example, if we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.26$ (from $\delta=0.1\pday$ and $\rw = 0\pday$).
Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
we do not claim that they are necessarily more accurate.
While individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
Instead, this calculation simply highlights the importance of carefully considering generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
We relied on parameter estimates for the Alpha variant throughout our analysis, but our qualitative conclusions can be applied to studying other SARS-CoV-2 variants, including the Delta variant.

@@

Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguments but sometimes gave inaccurate explanations. 
For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};
they hypothesized that a shorter generation interval of the Alpha variant could explain this phenomenon by reducing the relative speed (and therefore, the relative strength) under intervention measures.
However, our analysis suggests that a shorter generation interval of a new variant cannot explain the decline in relative strength.
Under a constant-strength intervention, the relative speed decreases (as predicted by  \cite{volz2021transmission}) but the relative strength remains constant (\fref{control_sim}B). 
Under a constant-speed intervention, the relative speed remains constant (\fref{control_sim}F), but the relative strength increases because the intervention will have a smaller effect on the variant with a shorter generation interval (\fref{control_sim}E).

@@

\bibliography{newvariant}

>>>>

Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
Analyses of new variants have primarily focused on characterizing changes in the \emph{proportion} of new variants, implicitly assuming that the relative speed remains fixed over the course of an invasion.
We use the generation-interval-based framework to challenge this assumption and illustrate how relative strength and speed change over time under two idealized interventions:
a constant-strength intervention like idealized vaccination and social distancing, which reduces transmission rates by a constant amount, and a constant-speed intervention like idealized contact tracing, which isolates infected individuals at a constant rate.
A constant-strength intervention causes relative speed to change, whereas a constant-speed intervention causes relative strength to change.
Differences in the generation-interval distributions between variants can exaggerate these changes and affect the effectiveness of interventions.
Neglecting the differences in the generation-interval distributions differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.

@@

Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution $g(\tau)$, an approach popularized by \citep{wallinga2007generation}.
The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
The generation interval is also distinctly different from other ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.

@@

Analyses of new variants have characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants, where $\Rv$, $\Rw$ and $\rv$, $\rw$ represent reproduction numbers and growth rates at a given time.
Many analyses have particularly focused on changes in the \emph{proportion} of a new variant to estimate the relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}, instead of taking the difference between their observed speed.
Focusing on proportions can be advantageous, because changes in proportions may be less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses
similarly;
however, estimates of relative speed from changes in the proportion of a new variant have typically relied on the assumption that the relative speed remains fixed over the time scale of an invasion.
Instead (or additionally), some studies have assumed a fixed value of the relative strength and tried to predict relative speed \citep{davies2021estimated,di2021impact}.

While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the pathogen and its control \citep{doi:10.1098/rspb.2020.1556}.
To illustrate the differences in implications for holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed in this paper.
Given a pre-intervention kernel $K(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals ($K(\tau)/\theta$)---in this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
On the other hand, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K(\tau) \exp(-\phi \tau)$---in this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to prevent onward transmission \citep{doi:10.1098/rspb.2020.1556}.

@@

Recent studies have further suggested the possibility that new variants may have different generation-interval distributions.
For example, Kissler \textit{et al.} \citep{kissler2021densely} suggested that the Alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the Alpha variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type.
Longer duration of infection suggests that the mean generation interval of the Alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
In contrast, some studies have suggested that the more recent Delta variant may have a shorter generation interval due to faster within-host viral replication \citep{li2021viral,zhang2021transmission}.
Modeling studies have also considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
However, linking strength and speed is complicated given that the generation-interval distribution depends on many factors including behavior:
for example, self-isolation after symptom onset will lead to shorter generation intervals.
The recent emergence of a new Omicron variant, and its breakthrough infections in previously immune individuals, adds further uncertainty to individual-level transmission dynamics and therefore the generation-interval of SARS-CoV-2 variants \citep{pulliam2021increased}.

@@

Under constant-strength changes that reduce the transmission rate by a constant amount across generation intervals, the instantaneous generation-interval distribution does not change through time \citep{fraser2007estimating}.
In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events \citep{champredon2015intrinsic}.
Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of constant-strength changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}---even though the instantaneous generation-interval distribution remains unchanged in this scenario.
While the instantaneous generation-interval distribution can change under speed-like changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that control strategies against its spread have been primarily strength-like, including social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
Indeed, many dynamical models of SARS-CoV-2 infections have solely relied on constant-strength changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution (e.g., \citep{flaxman2020Rt,gostic2020practical,brauner2021inferring}).
Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of constant-strength changes on the inference of dynamics of new SARS-CoV-2 variants.
We revisit these ideas in Section 6 and compare the effects of constant-speed interventions with those of constant-strength interventions.

@@

While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the proportion of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
\begin{align}
p(t) &= \frac{\iv(t_0) \exp(\rv t)}{\iw(t_0) \exp(\rw t) + \iv(t_0) \exp(\rv t)},
\\ &= \frac{1}{1 + \left(\iw(t_0)/\iv(t_0)\right) \exp(-\delta t)},
\end{align}
where the relative speed of the variant $\delta$ can be estimated as the slope of the log odds of $p$  vs.~time.
When more than two strains are co-circulating, the picture is more complicated \citep{campbell2021increased}; we focus here on comparing two strains at a time.

@@

% We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
Inference of the relative strength $\rho = \Rv/\Rw$ from the relative speed $\delta$ depends on assumptions about the generation-interval distributions of both strains.
Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
\begin{equation}
\rho = \left(\frac{1 + \kappa (\rw + \delta) \Gv}{1 + \kappa \rw \Gw}\right)^{1/\kappa}.
\end{equation}
Therefore, the relative strength $\rho$ depends not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population (\rw)---
some analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case, $\rho = \exp(\delta \Gw)$ when $\Gv=\Gw$).

@@

We do not generally expect the relative speed $\delta$ to remain fixed if other factors governing epidemic spread are changing on the time scale of the invasion of a new variant relative to that of a wild type.
Instead, many biological mechanisms appear compatible with assuming a fixed value of relative strength $\rho$ over changing conditions.
For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain unchanging, the relative strength $\rho$ is expected to remain fixed \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}; 
as we show in this section, these changes will imply a change in relative speed $\delta$.

We thus investigate how $\delta$ is expected to change with \Rw when $\rho$ remains fixed, and how this expectation changes with the ratio of the generation intervals. 
Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
\begin{equation}
\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \Gv} - \frac{\Rw^{\kappa} - 1}{\kappa \Gw}.
\end{equation}
As our baseline scenario we assume $\rho = 1.61$ (and therefore, we always have $\Rw < \Rv$), which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.

@@

\fref{relspeed} also shows that when $\rho$ is fixed, relative speed depends on underlying epidemiological conditions---specifically, the absolute strength of the two strains. 
For example, even when the generation-interval distributions are identical ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields a relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
Differences in the generation-interval distributions exaggerate these changes.
Therefore, characterizing changes in the proportion of variants by assuming a fixed relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant} and using the resulting value of $\delta$) should be done with care.

@@

Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the Alpha \citep{volz2021transmission} and Delta variants \citep{abbott2021estimating}.
Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}; 
therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission \citep{gostic2020practical}.
The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.

@@

We consider two types of intervention:
an intervention of constant strength (\fref{strengthspeed}A,C,E), which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed (\fref{strengthspeed}B,D,F), which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$);
both of these interventions are constant across generation intervals, but not necessarily across calendar time. 
In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively \citep{doi:10.1098/rspb.2020.1556}.
We consider constant-strength and -speed interventions calibrated to reduce $\Rv$ to 0.9 under the assumption that the variant generation intervals match those of the wild type (\fref{strengthspeed}C--D).
While both interventions are equally effective on the strength scale (that is, $\Ry{post}=\int  K_{\mathrm{post}}(\tau) \dtau = 0.9$), they have different dynamical implications.
The constant-strength intervention affects transmission equally throughout the course of infection, whereas the constant-speed intervention has greater impact on transmission that occurs later in infection;
as a result, the constant-speed intervention reduces the post-intervention mean generation interval (\fref{strengthspeed}B) and leads to (slightly) faster exponential decay (therefore, lower $\ry{post}$).

@@

We explored how the generation-interval distribution shapes the link between relative strength and speed.
While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, using the generation-interval-based framework provides clear explanations for an important, but under-appreciated concept: the relative strength and speed of new variants also depend on epidemic dynamics of the prior dominant lineages as well as the type of control measures (i.e., strength- or speed-like) that limit their spread.
For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain.
Our results challenge the commonly-made assumption that relative speed of new variants remains constant over time when characterizing changes in the proportion of new variants.
Differences in the generation-interval distributions further exaggerate this effect.

Therefore, neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the its transmission advantage.
These biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant.
Differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
If the variant has longer generation intervals than the wild type, speed-like interventions will be relatively more effective than naive estimates would suggest. 
Conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals.

There were a (minor) discrepancy in the estimates of relative strength of the Alpha variant.
Mathematical analyses typically reported greater than a 1.4 fold increase in reproduction number for the Alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggested a 1.25--1.4 fold increase \citep{ukinvest}.
As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences:
For example, if we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.26$ (from $\delta=0.1\pday$ and $\rw = 0\pday$).
Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
we do not claim that they are necessarily more accurate.
While individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
Instead, this calculation simply highlights the importance of carefully considering generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
We relied on parameter estimates for the Alpha variant throughout our analysis, but our qualitative conclusions can be applied to studying other SARS-CoV-2 variants, including the Delta and Omicron  variant.

Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguments. 
For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};
they hypothesized that a shorter generation interval of the Alpha variant could explain this phenomenon by reducing the relative speed (and therefore, the relative strength) under intervention measures.
However, our analysis suggests that a shorter generation interval of a new variant cannot explain the decline in relative strength.
Under a constant-strength intervention, the relative speed decreases (as predicted by  \cite{volz2021transmission}) but the relative strength remains constant (\fref{control_sim}B). 
Under a constant-speed intervention, the relative speed remains constant (\fref{control_sim}F), but the relative strength increases because the intervention will have a smaller effect on the variant with a shorter generation interval (\fref{control_sim}E).

We have primarily focused on the differeces in the mean generation interval but the differences in the amount of variability, characeterized by the squared coefficient of variation $\kappa$, can also have important, but different implications \citep{blanquart2021selection}.
For example, if two variants have identical reproduction numbers, the variant with a wider generation-interval distribution will always have a faster growth rate and outcompete the other variant (except when both variants have $\RR=1$).
This is because a variant with a wider generation-interval distribution can take advantage of more early transmission during the growth phase and more late transmission during the decay phase.s

@@

@@

\bibliography{newvariant_abbv}
